Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions

### 2025-08-17 paper.qmd
Making changes

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail index directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, discrete extremes,power law, preferential attachment]
bibliography: refs.bib  
---

@@

Networks appear in many fields such as sociology, politics, epidemiology, and economics. Statistical methods have been used to study networks that appear in these fields and beyond, from the use of stochastic block models to detect communities within the French political blogosphere [@Latouche11], to using Exponential Random Graph Models (ERGMs) to analyse the global trade network [@Setayesh22], and the use of mechanistic models in neuroscience to investigate the patterns of connections in neural systems [@Betzel17].

One example of mechanistic models is the preferential attachment (PA) model, where as a network grows and new vertices join, they connect to those already present in the network with probabilities proportional to some preference function of their in-degrees. The well-known Barabási-Albert (BA) model is a special case of this when $b(k) = k+1$ [@Barabasi99].

These mechanistic models are generally useful when simulating networks, but not for fitting to real data as usually only snapshots of a network are available and not the full evolution - without this conclusions about the growth are difficult to make. However, one statistic that is usually available for snapshots of networks is the degree distribution, making it valuable to study the degrees in a network. Often, it is appealing to model the degree distributions of real networks using the power law [^1], as it is seemingly observed in much of the data. Furthermore, as the BA model results in a power law degree distribution [@Barabasi99], that the power law fits well offers the interpretation that the network may have grown according to PA or similar simple rules. However, attempting to explain the networks growth using preferential attachment when the degrees seemingly follow a power law may not be the best course of action as @krapivsky01 show that a general preference function does not necessarily lead to power law degrees. Specifically, showing that a preference function of form $b(k) = k^\alpha$ leads to light tails when $\alpha<1$, a degenerate network (a finite number of vertices end up with all of the added edges) when $\alpha>1$, and only yields heavy tails when $\alpha=1$. @rudas07 derives the limiting degree distribution in terms of the preference function $b(\cdot)$ and its parameters, we build upon this by finding a class of preference functions that result in realistic tails i.e. heavy but not as heavy as implied by the power law. On the other hand, it is still being argued whether or not many real data actually follow the power law. @Broido_2019 provides some evidence that real world networks rarely have degrees that follow a power law, through the use of statistical analysis of almost 1000 networks across a wide variety of fields; @Voitalov_2019, on the other hand, disagrees and claims that these kinds of networks are not nearly as rare and only appear so as a result of an unrealistic expectation of power law without deviations or noise.

[^1]: Power law: $\bar F(k) = L(k)\times k^{-1/\xi}$, where $\xi>0$ and $L(k)$ is slowly varying in $k$.

Extreme value methods have been scarcely used when it comes to analysis of degree distributions and whether they follow the power law. The absence of these methods essentially downweights the deviation of vertices with the largest degrees, which are usually the most influential, from the power law. @Voitalov_2019 proposes estimating the power law exponent of a degree distribution using three estimators based in extreme value theory, in contrast to the estimators generally used in network science, the degree distributions are then split into one of three categories (not power-law, hardly power-law or power-law) depending on the values of the estimates obtained from each of the estimators. @Lee24 proposes a framework for modelling entire degree distributions using a combination of two generalisations of the power law with a model selection step that is capable of suggesting whether or not the data is adequately modelled by a power law, with one conclusion being that most data seems to have a tail lighter than what is implied by the power law in body whilst still being heavy tailed. One of the models used by @Lee24 is the Zipf-IGP model where up until a certain threshold the data is modelled using the Zipf disribution (a discrete power law) and beyond it a discrete version of the Generalised Pareto distribution is used, we will later compare our model to this as they share some similarities. @wang2022random considers a preferential attachment model with an affine preference function and reciprocal edges, finding that the joint limiting distribution of the in-degrees and out-degrees is multivariate regularly varying and has the property of hidden regular variation.

Evidently, the study of degree distributions is a very active but the vast majority of research in the field (whether using extremes or not) are purely descriptive in the sense that no information about the preference function is revealed even if preferential attachment was the underlying mechanism. This paper aims to address this gap, asking if the degree distribution of a network is assumed to come from the preferential attachment model, can we directly infer the model parameters?

The remainder of this paper is as follows: Section 2 gives a detailed description of the PA model alongside the theoretical results for the limiting degree distribution with a focus on its tail behaviour in terms of the preference function $b(\cdot)$. It is shown that a heavy tailed degree distribution can only be the result of an asymptotically linear preference function ending with a suggestion for a preference function that can guarantee heavy tails while remaining flexible. @sec-model utilises the preference function proposed at the end of Section 2, and illustrates how the tail index of the degree distribution varies with the model parameters. @sec-sim consists of a simulation study where networks are simulated from the proposed model for various parameter combinations, demonstrating that the parameters can be recovered from fitting the model to only the degree distribution. @sec-real fits the model to some real data and provides posterior estimates for the preference function. @sec-conc concludes the article.

@@

The network generative model that we will focus on in this paper is the General Preferential Attachment model in @rudas07 and is defined as follows:

Starting at time $t=0$ with an initial network of $m$ vertices that each have no edges, at times $t=1,2,\ldots$ a new vertex is added to the network bringing with it $m$ directed edges (with the new vertex as the source); the target for each of these edges are selected from the vertices already in the network with weights proportional to some preference function $b(\cdot)$ of their degree, where $b: N \mapsto \mathbb R^+\setminus\{0\}$ is such that:

@@

Special cases of this model include the Barabási-Albert (BA) model when $b(k) = k+\varepsilon$, which leads to a power-law degree distribution with $\xi=1/2$, and the Uniform Attachment (UA) model where $b(k)=c$ leading to a degree distribution that is light tailed.

Given condition [-@eq-condb2] the survival function of the limiting degree distribution, called the limiting survival hereafter, can be analytically derived in the case where $m=1$, obtained by considering a branching process that is equivalent to the growth of the network, as in @rudas07. Theorem 1 from @rudas07 states that for the tree $\Upsilon(t)$ at time $t$:

$$
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
$$ {#eq-survlim} where $\lambda^*$ satisfies $\hat\rho(\lambda^*)=1$ and $\hat\rho$ is the Laplace transform of the density of the point process associated with the pure birth process that corresponds to a nodes individual growth, that is:

$$
\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t.
$$

@@

$$
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
$$ Therefore, the corresponding probability mass function of the degree distribution $f(k) = \bar F(k-1) - \bar F(k)$ is

$$
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)}.
$$

We are interested in how the tail behaviour of the discrete limiting degree distribution is affected by the preference function $b$, which can be determined by the results of @shimura12.

For a distribution $F$ with survival function $\bar F$ and some $k\in\mathbb Z^+$ let

@@

@shimura12 states that if $\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha$ ($\alpha>0$), then $F$ is heavy-tailed with $\bar F(k) \sim k^{-\alpha}$. On the other hand, if $\lim_{k\rightarrow\infty} \Omega(F,k) = 0$ then the distribution is light-tailed. This allows us to show the following:

@@

@prp-omega aligns with the result from @krapivsky01 demonstrating that a sub-linear preference function will lead to a light-tailed distribution, as $\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0$ if $b(k)=k^\alpha$ where $\alpha < 1$. Proposition 2.1 also aligns with the fact that BA model produces a heavy-tailed degree distribution with index $\xi=0.5$ by considering the preference function $b(k) = k + \varepsilon$, as $\lim_{k\rightarrow\infty}b(k+1)-b(k)=1$ leaving the tail index to be $1/\lambda^*$ which using $\hat\rho$ can be found to be $1/2$. So, in order for the degree distribution to be heavy-tailed we need that the limit $\lim_{k\rightarrow\infty} b(k+1)-b(k)$ exists and is positive. To determine the class of functions that will result in heavy-tailed limiting degree distributions, we use the following result:

@@

Consider a GPA model with preference function $b(\cdot)$ satisfying @prp-omega. Then the limiting degree distribution has a heavy tail with index $c/\lambda^{*}$ if and only if $\lim_{k\rightarrow\infty}b(k)/k = c > 0$.

@@

$$
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
$$ meaning by @thm-stolz:

@@

Considering @prp-omega2 we propose the following preference function:

@@

This preference function, as per @prp-omega, will produce a degree distribution with tail index $\beta/\lambda^*$ guaranteeing a heavy tail. We study this preference function further in the next section.

@@

# Preferential Attachment with Flexible Heavy Tail {#sec-model}

In the previous section, we found that using a preference function with linearity in the limit allows for the inclusion of sub/super-linear behaviour below the threshold, while simultaneously guaranteeing a heavy tail. In this section, we study how the limiting degree distribution varies with the preference function according to @eq-pref. Its survival is

@@

which must be solved numerically for most parameter choices. Also, note that $\lambda>\beta$.

Some examples of the limiting degree distribution for various parameter combinations are shown below on log-log scale in @fig-polylinsurv:

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: Theoretical survival distributions of the limiting degree distributions, according to various combinations of $(\alpha, \beta, \varepsilon)$ and $k_0=20$ of the proposed preferential attachment model.
#| label: fig-polylinsurv

@@

pars[which.max(pars$b/lambdas),]
max(pars$b/lambdas)

ggplot(data = pars) + geom_line(aes(x=(x+1),y=surv, linetype=as.character(b),colour =as.character(b))) +
  scale_x_log10() + scale_y_log10(limits=c(1e-5,1))+theme(aspect.ratio = 1/2) + theme_bw() + xlab('Total Degree')  +ylab('Survival') + labs(linetype=TeX('\\beta'), colour =TeX('\\beta')) + 
  facet_grid(eps~a,labeller = labeller, scales='free')
```
@fig-polylinsurv demonstrates that this model can capture a wide variety of shapes for the survival function, including a large range of possible tail indices ranging from  0.035 ($\alpha=1.5, \beta=0.1, \varepsilon=1$) to 0.999  ($\alpha=0.5, \beta=1.5, \varepsilon=0.01$).

The survival function  (-@eq-polysurv) can be connected to the discrete version of the generalised Pareto distribution (GP), called the Integer GP (IGP) [@Rohrbeck_2018] with conditional survival:
$$
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
$$ for $v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R$, denoted as $X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)$ where $\xi$ is the shape parameter controlling the tail index

@@

\begin{align*}
\bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\
&\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\
&=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\
&=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta}
\end{align*}

Therefore, 
$$
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
$$ meaning that for $k\ge k_0$ the limiting degree distribution (for large $k_0^\alpha$) is approximated by $\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)$.

To assess how close of an approximation this is, the theoretical conditional survivals are shown in @fig-approx_surv in colour and their IGP approximations are shown in grey. The approximation seems to hold up fairly well even for large degrees and more so when $\alpha$ is larger.

@@

In agreement with @prp-omega2, $\beta>0$ ensures that the shape parameter of the IGP is positive and thus the distribution is heavy-tailed. Additionally the value of the shape parameter $\xi$ is shown in @fig-polyheat for various parameter choices. The darker and lighter regions on the heat maps correspond to a heavier and a lighter tail, respectively, the red dashed line shows combinations of $\alpha$ and $\beta$ that produce a limiting degree distribution with the same tail index as the Barabási-Albert model, that is, $\xi=0.5$.

@@

To perform inference of the model parameters, we consider a network with degree count vector $\pmb n = (n_0, n_1, \ldots, n_M)$, where $M$ is the maximum degree. We can then write the likelihood as:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}\label{eq-lh}

>>>>

---
title: "Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions"
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail index directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, discrete extremes,power law, preferential attachment]
bibliography: refs.bib  
keep-tex: true
pdf-engine-opt: -outdir=''
output-temp-dir: 'temporary_files'
---

@@

Networks have become very powerful tools for representing and analysing complex systems, with uses in a large array of fields. In network science and statistics, they have been studies by various families of models, from stochastic block models for detecting communities online [@Latouche11], to exponential random graph models (ERGMs) for analysing the global trade network [@Setayesh22],  and mechanistic models for investigating patterns in neural systems [@Betzel17].

Amid the recent rise of interest in networks, there has been a debate on whether most real networks are scale-free. Claiming a real network is scale-free is equivalent to saying that its degree distribution follows a power law, that is the fraction of nodes with degree $k$ is proportional to $k^{-\alpha}$, and therefore has a regularly varying tail with extreme value index $1/\alpha$. On the side against the claim is  @Broido_2019who compared the fits of a power law model against that of several non-scale-free models to nearly a thousand networks, only to find strong evidence for scale-freeness in four percent and weak evidence in over half of the networks, thus claiming that scale-free networks do not make up a majority in real networks. They compare the fits of a power law model against that of several non-scale-free models only finding strong evidence for scale-freeness in four percent and weak evidence in fifty-two percent of the networks. On the other side of this debate is @Voitalov_2019 who disagrees and claims that these networks are not nearly as rare and only appear to so as a result of an unrealistic expectation of a power law without deviations or noise. Additional evidence of these deviations from a power law is shown by @Lee24 who demonstrate that a lot of networks are partially scale-free, in that the body of the degree distribution is often modelled well by a power law, while the tail is lighter than what is implied by the body, albeit still regularly varying..

Nevertheless, most studies into the appropriateness of a power law for the degrees of real networks, the aforementioned references included, have been largely descriptive in the sense that no information about the growth of the networks is revealed.

The popularity of using the power law for network degrees can be traced back to the preferential attachment (PA) model popularised by @Barabasi99. In the general model, as new nodes join the network, an existing node with degree k gains edges at a rate proportional to $b(k)$, where b is a non-negative preference function. @Barabasi99 showed that, when $b(k) = k + 1$, in the limit the resulting degree distribution is regularly varying with index 3. Subsequently, if a real network is shown to be scale-free, one can loosely justify preferential attachment as the underlying mechanism of its growth.

The model from @Barabasi99 provided the foundations for various generalisations.

@krapivsky01 considered $b(k) = (k+1)^\alpha$, and showed that the degree distribution is not regularly varying (and therefore not following the power law) when $0<\alpha<1$, and when $\alpha>1$ a finite number of nodes end up with all edges after a certain point resulting in a degenerate degree distribution. @wang2022random returns to a linear preference function of the form $b(k) = k+\varepsilon$ but adds the possibility for reciprocal edges to be sent, resulting in the joint distribution of in-degree and out-degree being multivariate regularly varying and having the property of hidden regular variation. @rudas07 follows in the footsteps of @krapivsky01 , by considering a preferential attachment tree and using theory from continuous branching processes, derives a limiting degree distribution in terms of the preference function $b(\cdot)$. Nevertheless, research in this area tends to only focus on the theoretical asymptotic results of network growth models with little analysis of real networks.

This paper aims to address the gap between the applied and theoretical works, by asking if a network is assumed to come from a preferential attachment model, can we use the degree distribution alone to directly infer the parameters of the the preference function and learn about the growth mechanisms? Moreover, proper consideration is given to the tail of the degree distribution, because otherwise the effects of the largest degrees, which correspond to the most influential nodes, deviating from the power law will be discounted.

It is important to think about how we intend to consider the tail of the degree distributions, because if not done correctly we may end up essentially discounting the effects of the largest degrees (often the most influential nodes) deviating from a power law.

For the above reason, we will use methods from discrete extreme value theory, in particular those by @shimura12, who provided theoretical guarantees for a discrete distribution to be regularly varying or not. Using these results, we demonstrate how the tail of the degree distribution is affected by $b(\cdot)$, and subsequently propose a class of preference functions that is tail realistic for real networks. These analytical results enables the likelihood of the degree distribution to be expressed in terms of the parameters of $b(\cdot)$, which in turn allows the underlying mechanism of the network, assumed to grow according to preferential attachment, to be inferred directly.

The remainder of the paper is as follows: @sec-tail gives a detailed description of the preferential attachment model alongside the theoretical results for the survival of the limiting degrees, with a focus on the tail behaviour in terms of the preference function $b(\cdot)$. A class of asymptotically linear preference functions will be introduced and shown to guarantee regular variation in the degrees while remaining flexible up until a threshold. @sec-model utilises the proposed preference function and illustrates numerically how the extreme value index (EVI) of the degree distribution varies with the model parameters. @sec-sim consists of a simulation study where networks are simulated from the proposed model for various parameter combinations, demonstrating that the parameters can be recovered from fitting the model to only the degree distribution. @sec-real fits the model to some real data and provides posterior estimates for the preference function. @sec-conc concludes the article.

@@

The model that we will focus on in this paper is the General Preferential Attachment model in @rudas07 and is defined as follows:

Starting at time $t=0$ with an initial network of $m$ vertices that each have no edges, at times $t=1,2,\ldots$ a new vertex is added to the network bringing with it $m$ directed edges from the new vertex; the target for each of these edges are selected from the vertices already in the network with weights proportional to some preference function $b(\cdot)$ of their degree, where $b: \mathbb N \mapsto \mathbb R^+\setminus\{0\}$ is such that:

@@

Special cases of this model include the Barabási-Albert (BA) model when $b(k) = k+\varepsilon$, which leads to a power-law degree distribution with  EVI $\xi=1/2$, and the Uniform Attachment (UA) model where $b(k)=c$ leading to a degree distribution that is not regularly varying.

The survival function of the limiting degree distribution, called the limiting survival hereafter, under condition [-@eq-condb2] can be analytically derived in the case where $m=1$, this is presented below.

Consider a continuous time branching process $\zeta(t)$ driven by a Markovian pure birth process, with $\zeta(0)=0$ and birth rates depending on a non-negative function $b(\cdot)$:

$$
\Pr(\zeta(t+dt)=k+1|\zeta(t)=k) = b(k)dt + o(dt).
$$
Now let $\Upsilon(t)$ be the tree determined by $\zeta(t)$ as follows: $\Upsilon(t)=\{\emptyset\}$ and $\Upsilon(t)=G$ where each existing node $x$ in $\Upsilon(t)$ gives birth to a child with rate $b(\mathrm{deg}(x, \Upsilon(t))$ independently of the other nodes where $\mathrm{deg}(x, \Upsilon(t))$ is the degree of node $x$ in the tree $\Upsilon(t)$ at time $t$.

Theorem 1 from @rudas07 states that for the tree $\Upsilon(t)$ at time $t$:

$$
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
$$ {#eq-survlim} 
where $\lambda^*$ satisfies $\hat\rho(\lambda^*)=1$ and $\hat\rho$ is the Laplace transform of the density of the point process associated with the pure birth process that corresponds to the growth of an individual node, that is $\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t$

@@

$$
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
$${#eq-surv-origin}

Therefore, the corresponding probability mass function of the degree distribution $f(k) = \bar F(k-1) - \bar F(k)$ is

$$
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)}.
$${#eq-pmf-origin}

We are interested in how the tail behaviour of the discrete limiting degree distribution is affected by the preference function $b$.

For a distribution $F$ with survival function $\bar F$ and $k\in\mathbb Z^+$ let

@@

This allows us to show the following:

@@

@shimura12 states that if $\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha$ ($\alpha>0$), then $F$ is regularly varying with $\bar F(k) \sim k^{-\alpha}$. On the other hand, if $\lim_{k\rightarrow\infty} \Omega(F,k) = 0$ then we will refer to the distribution as light-tailed.

@prp-omega aligns with the result from @krapivsky01 demonstrating that a sub-linear preference function will lead to a light-tailed distribution, as $\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0$ if $b(k)=k^\alpha$ where $\alpha < 1$. Proposition 2.1 also aligns with the fact that BA model produces a regularly varying degree distribution with EVI $\xi=0.5$ by considering the preference function $b(k) = k + \varepsilon$, as $\lim_{k\rightarrow\infty}b(k+1)-b(k)=1$ leaving the tail index to be $1/\lambda^*$ which using $\hat\rho$ can be found to be $1/2$. So, in order for the degree distribution to be regularly varying  we need that the limit $\lim_{k\rightarrow\infty} b(k+1)-b(k)$ exists and is positive. To determine the class of functions that will result in regularly varying limiting degree distributions, we use the following result:

@@

Consider a GPA model with preference function $b(\cdot)$ satisfying @prp-omega. Then the limiting degree distribution is regularly varying with EVI $c/\lambda^{*}$ if and only if $\lim_{k\rightarrow\infty}b(k)/k = c > 0$.

@@

$$
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
$$ Using the Stolz-Cesaro Theorem [@thm-stolz](Appendix -@sec-sup),

@@

Using this result we can understand how the preference function is directly connected with the tail behaviour of the degree distribution. Specifically, regular variation is achieved if and only if $b(k)$ is asymptotically linear with $k$. We use this result to create a preference function that guarantees regular variation in the tail of the degree distribution, aligning with analysis of real networks, whilst allowing for the tail to deviate from the shape implied by the body. This gives the model the capability to produce realistic behaviour in the degrees like what was in @Lee24 by using a piecewise function inspired by the observed deviation from the power law after a certain threshold:

@@

Using @prp-omega, we can show that $\lim_{n\rightarrow\infty}\Omega(F,k)=\lambda^{*}/\beta$, meaing the degree distribution obtained using this preference funnction is regularly varying with tail index $\lambda^*/\beta$.

@@

# A Preferential Attachment Model with Flexible Regular Variation {#sec-model}

In the previous section, we found that using an asymptotically linear preference function allows for the inclusion of sub/super-linear behaviour below the threshold, while simultaneously guaranteeing regular varation of the degrees. In this section, we demonstrate the flexibility of the preference function in @eq-pref, with regard to the tail behaviour of the limiting degree distribution. Using @eq-surv-origin, the limiting survival is

@@

which has to be solved numerically for most parameter choices. Also, note that $\lambda>\beta$.

For some parameter combinations, the limiting survival $\bar F(k)$ is shown on log-log scale in @fig-polylinsurv:

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: The limiting survival, according to various combinations of $(\alpha, \beta, \varepsilon)$ and $k_0=20$ of the proposed preferential attachment model.
#| label: fig-polylinsurv

@@

# pars[which.max(pars$b/lambdas),]
# max(pars$b/lambdas)

ggplot(data = pars) + geom_line(aes(x=(x+1),y=surv, linetype=as.character(b),colour =as.character(b))) +
  scale_x_log10() + scale_y_log10(limits=c(1e-5,1))+theme(aspect.ratio = 1/2) + theme_bw() + xlab('Total Degree')  +ylab('Survival') + labs(linetype=TeX('\\beta'), colour =TeX('\\beta')) + 
  facet_grid(eps~a,labeller = labeller, scales='free')
```
@fig-polylinsurv demonstrates that this model can capture a range of tail behaviour, including a large range of possible tail indices ranging from  0.035 ($\alpha=1.5, \beta=0.1, \varepsilon=1$) to 0.999  ($\alpha=0.5, \beta=1.5, \varepsilon=0.01$).

The analytic form of the survival function  in (-@eq-polysurv), offers a natural connection to the discrete version of the generalised Pareto distribution (GP), providing a link to a well established component of discrete extremes in the literature. Specifically, connecting to the Integer GP (IGP) seen in @Rohrbeck_2018 with conditional survival:
$$
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
$$ for $v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R$, denoted as $X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)$ where $\xi$ is the shape parameter controlling the tail index.

@@

\begin{align*}
\bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\
&\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\
&=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\
&=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta}.
\end{align*}

Therefore, 
$${#eq-igp-est}
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
$$ meaning that for $k\ge k_0$ the limiting degree distribution (for large $k_0^\alpha$) is approximated by the $\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)$.

To assess how close of an approximation this is, the theoretical conditional survival [@eq-polysurv] are shown in @fig-approx_surv in colour and their IGP approximations [@eq-igp-est] are shown in grey. The approximation seems to hold up fairly well even for large degrees and more so when $\alpha$ is larger.

@@

In agreement with @prp-omega2, $\beta>0$ ensures that the shape parameter of the IGP is positive and thus the distribution is regularly varying. Additionally the value of the shape parameter $\xi$ is shown in @fig-polyheat for various parameter choices. The darker and lighter regions on the heat maps correspond to a heavier and a lighter tail, respectively, the red dashed line shows combinations of $\alpha$ and $\beta$ that produce a limiting degree distribution with the same tail index as the Barabási-Albert model, that is, $\xi=0.5$.

Through the connection to the IGP, fitting this model is almost equivalent to fitting the IGP to the degrees and estimating its parameters. However, instead of only describing the shape of the degree distribution, we would also gain estimates for the shape of the preference function, thus gaining a direct understanding the mechanisms underlying the growth of the network.

@@

To perform inference of the model parameters, we consider a network with degree count vector $\pmb n = (n_0, n_1, \ldots, n_M)$, where $M$ is the maximum degree. Using @eq-pmf-origin, the likelihood is:

\begin{align*}
L(\pmb n | \pmb \theta,l) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}\label{eq-lh}
